---
title: "Data Cleaning, EDA, and Modeling"
author: "Jaret Stickrod"
output: html_document
date: "2024-11-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Install required libraries if not already installed
if (!requireNamespace("mgcv", quietly = TRUE)) install.packages("mgcv")
if (!requireNamespace("caret", quietly = TRUE)) install.packages("caret")
if (!requireNamespace("tidyverse", quietly = TRUE)) install.packages("tidyverse")
if (!requireNamespace("dplyr", quietly = TRUE)) install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE)) install.packages("ggplot2")
if (!requireNamespace("randomForest", quietly = TRUE)) install.packages("randomForest")
if (!requireNamespace("xgboost", quietly = TRUE)) install.packages("xgboost")

```


### Research Question: How does the number of defenders in the box and number of pass rushers affect the play result?

## Data Cleaning

```{r}
library(dplyr)
library(tidyverse)

df <- read.csv("plays.csv")

# Filter for passing plays only
df <- df %>%
  filter(playType == "play_type_pass") %>%
  select(playResult, defendersInTheBox, numberOfPassRushers) %>%
  drop_na()

# Binning the predictors
df <- df %>%
  mutate(
    defendersInTheBox_binned = cut(defendersInTheBox, breaks = seq(4, 8, by = 1)),
    numberOfPassRushers_binned = cut(numberOfPassRushers, breaks = seq(3, 7, by = 1))
  )
str(df)
```


## Exploratory Data Analysis

```{r}
##### Boxplots

library(ggplot2)

ggplot(df, aes(x = factor(defendersInTheBox), y = playResult)) +
  geom_boxplot() +
  labs(x = "Defenders in the Box", y = "Play Result", title = "Impact of Defenders on Play Result")

ggplot(df, aes(x = factor(numberOfPassRushers), y = playResult)) +
  geom_boxplot() +
  labs(x = "Number of Pass Rushers", y = "Play Result", title = "Impact of Pass Rushers on Play Result")

##### Heatmap
heatmap_data <- df %>%
  group_by(defendersInTheBox, numberOfPassRushers) %>%
  summarise(mean_playResult = mean(playResult, na.rm = TRUE))

ggplot(heatmap_data, aes(x = defendersInTheBox, y = numberOfPassRushers, fill = mean_playResult)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  labs(title = "Heatmap of Play Result by Defenders and Pass Rushers", fill = "Mean Play Result")

```
```{r}
##### Check correlations using correlation matrix
numeric_vars <- c("playResult", "defendersInTheBox", "numberOfPassRushers")
cor_matrix <- cor(df[, numeric_vars], use = "complete.obs")
print(cor_matrix)

```



## Exploratory Modeling 

```{r}
library(dplyr)
library(mgcv)
library(caret)

##### Splitting the Data
set.seed(123)  # reproducibility
train_index <- createDataPartition(df$playResult, p = 0.8, list = FALSE)
train_data <- df[train_index, ]
test_data <- df[-train_index, ]

##### Exploratory Modeling
# Linear Regression
lm_model <- lm(playResult ~ defendersInTheBox + numberOfPassRushers, data = train_data)

# Generalized Additive Model (GAM)
gam_model <- gam(playResult ~ s(defendersInTheBox) + s(numberOfPassRushers), data = train_data)

##### Cross-Validation
set.seed(445)
control <- trainControl(method = "cv", number = 5)  # 5-fold cross-validation

# Cross-Validation for Linear Regression
cv_lm <- train(
  playResult ~ defendersInTheBox + numberOfPassRushers,
  data = train_data,
  method = "lm",
  trControl = control
)

# Cross-Validation for GAM
cv_gam <- train(
  playResult ~ defendersInTheBox + numberOfPassRushers,
  data = train_data,
  method = "gamSpline",
  trControl = control
)

##### Model Performance on Test Data
# Predict and Evaluate Linear Regression
lm_predictions <- predict(lm_model, newdata = test_data)
lm_rmse <- sqrt(mean((lm_predictions - test_data$playResult)^2))

# Predict and Evaluate GAM
gam_predictions <- predict(gam_model, newdata = test_data)
gam_rmse <- sqrt(mean((gam_predictions - test_data$playResult)^2))

# Output Results
cat("=== Model Summaries ===\n")
cat("Linear Regression:\n")
print(summary(lm_model))
cat("\nGAM:\n")
print(summary(gam_model))

cat("\n=== Cross-Validation Results ===\n")
cat("Linear Regression Cross-Validation:\n")
print(cv_lm)
cat("\nGAM Cross-Validation:\n")
print(cv_gam)

cat("\n=== Test Set Performance ===\n")
cat("Linear Regression RMSE:", lm_rmse, "\n")
cat("GAM RMSE:", gam_rmse, "\n")
```

Model Coefficients (Play Result outcome):

Linear Regression: 
                    Estimate Std. Error t value Pr(>|t|)    
(Intercept)          5.78934    0.59994   9.650  < 2e-16 ***
defendersInTheBox    0.00161    0.08767   0.018 0.985344    
numberOfPassRushers  0.38050    0.10086   3.773 0.000162 ***

GAM:
                         edf Ref.df     F  p-value    
s(defendersInTheBox)   4.436  5.368 4.308 0.000498 ***
s(numberOfPassRushers) 2.968  3.760 7.828 6.14e-06 ***


1. In General, both linear regression and GAM explain very little variance in `playResult`, with low R-squared values and high RMSE:
Linear Regression RMSE: 9.847673 
GAM RMSE: 9.834928

2. The number of pass rushers shows a stronger relationship with `playResult` (P-value = 0.000162), compared to defenders in the box (P-value = 0.985).  

3. The GAM slightly outperforms linear regression, but the improvement is not significant.  
 
 
## Visualizing models above
```{r}
library(ggplot2)
library(mgcv)
library(dplyr)

# Generate predictions for both models
train_data <- train_data %>%
  mutate(
    lm_predictions = predict(lm_model, newdata = train_data),
    gam_predictions = predict(gam_model, newdata = train_data)
  )

# Visualization for Linear Regression
ggplot(train_data, aes(x = defendersInTheBox, y = playResult)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_line(aes(y = lm_predictions), color = "red", size = 1) +
  labs(
    title = "Linear Regression Model",
    x = "Defenders in the Box",
    y = "Play Result"
  ) +
  theme_minimal()

# Visualization for GAM - Defenders in the Box
plot(gam_model, select = 1, se = TRUE, rug = TRUE, main = "GAM: Effect of Defenders in the Box")

# Visualization for GAM - Number of Pass Rushers
plot(gam_model, select = 2, se = TRUE, rug = TRUE, main = "GAM: Effect of Number of Pass Rushers")

# Combined Visualization for GAM (Optional)
gam_predictions <- data.frame(
  defendersInTheBox = train_data$defendersInTheBox,
  numberOfPassRushers = train_data$numberOfPassRushers,
  gam_predictions = train_data$gam_predictions
)

ggplot(train_data, aes(x = defendersInTheBox, y = playResult)) +
  geom_point(alpha = 0.5, color = "blue") +
  geom_line(data = gam_predictions, aes(y = gam_predictions), color = "green", size = 1) +
  labs(
    title = "GAM Model",
    x = "Defenders in the Box",
    y = "Play Result"
  ) +
  theme_minimal()

```



### Secondary Research Question: Can we classify plays (playType) into sacks, completed passes, or interceptions based on the number of defenders in the box, number of pass rushers, and other features?

```{r}
#####  Logistic regression to predict likelihood of a sack

library(dplyr)
library(tidyverse)

playTypedf <- read.csv("plays.csv")

df <- playTypedf %>%
  select(playType, defendersInTheBox, numberOfPassRushers, down) %>%
  drop_na()


logistic_model <- glm(I(playType == "sack") ~ numberOfPassRushers + down,
                      family = binomial, data = df)
summary(logistic_model)

```

```{r}
library(randomForest)
library(caret)

clean_df <- na.omit(df[, c("defendersInTheBox", "playResult", "numberOfPassRushers")])

set.seed(123)
train <- createDataPartition(clean_df$playResult, p = 0.8, list = FALSE)
train_data <- clean_df[train, ]
test_data <- clean_df[-train, ]

rf <- randomForest(playResult ~ defendersInTheBox + numberOfPassRushers, data = train_data, ntree = 100)

prediction <- predict(rf, test_data)

actual <- factor(round(test_data$playResult))
predicted <- factor(round(prediction), levels = levels(actual))

cm <- confusionMatrix(predicted, actual)
cm
importance(rf)
varImpPlot(rf)
```
The random forest is not a good model to use given it can only accurately predict the relationship
4.53% of the time. Since numberOfPassRushers has an importance score of 6791 and defendersInTheBox has an importance score of 4881 the model is suggesting that numberOfPassRushers is more influential in predicting the outcome of a play. 

```{r}
library(xgboost)
library(dplyr)

set.seed(123)

response <- "playResult"
predictors <- c("defendersInTheBox", "numberOfPassRushers")

plays <- df %>%
  mutate(across(all_of(c(response, predictors)), as.numeric))

train <- createDataPartition(plays[[response]], p = 0.8, list = FALSE)
train_data <- plays[train, ]
test_data <- plays[-train, ]

train_x <- as.matrix(train_data[, predictors])
train_y <- train_data[[response]]

test_x <- as.matrix(test_data[, predictors])
test_y <- test_data[[response]]

params <- list(
  bjective = "reg:squarederror", 
  eta = 0.1,      
  max_depth = 6,   
  subsample = 0.8, 
  colsample_bytree = 0.8
)

xgb <- xgboost(
  data = train_x,
  label = train_y,
  params= params,
  nrounds = 100,
  verbose = 0
)

xgb_predict <- predict(xgb, test_x)

xgb_mse <- mean((test_y - xgb_predict)^2)
xgb_rmse <- sqrt(xgb_mse)

xgb_rmse
importance <- xgb.importance(feature_names = predictors, model = xgb)
xgb.plot.importance(importance)
```

```{r}
library(pdp)

train_data <- data.frame(defendersInTheBox = rnorm(100), 
                         numberOfPassRushers = rnorm(100), 
                         playResult = rnorm(100))

xgb_model <- xgboost(data = as.matrix(train_data[, c("defendersInTheBox", "numberOfPassRushers")]),
                     label = train_data$playResult,
                     nrounds = 100,
                     objective = "reg:squarederror")

pdp_defenders <- partial(xgb_model, pred.var = "defendersInTheBox", train = train_data)

plot(pdp_defenders)
##Still Debugging, shows how feature affects target holding other variables constant 
```

```{r}
library(e1071)  
library(caret)  

svm_model <- svm(playResult ~ defendersInTheBox, 
                 data = train_data, 
                 type = "eps-regression",  
                 kernel = "radial")       

predictions <- predict(svm_model, newdata = test_data)

if (is.numeric(test_data$playResult)) {
  rmse <- sqrt(mean((predictions - test_data$playResult)^2))
  cat("RMSE:", rmse, "\n")
} else {
  confusion <- confusionMatrix(as.factor(predictions), as.factor(test_data$playResult))
  print(confusion)
}

```