---
title: "NFL Big Data Bowl 2021"
author: "Ty Hammond, Blake Hammarstrom, Jaret Stickrod, and Luke Spencer"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Motivation and Design

-   **Main Point:** Understanding how defensive strategies impact NFL play outcomes.

    -   **Focus:** Investigate if more defenders in the box reduce offensive yards gained.

    -   **Significance:**

        -   Helps coaches balance between rushing and coverage plays.

        -   Guides decisions on optimizing defensive play-calling.

        -   Example:

            -   More pass rushers = quarterback pressure but open downfield.

            -   Fewer rushers = time for quarterback but vulnerable coverage gaps.

    -   **Inspiration:**

        -   Shared passion for football and analytics.

        -   Contribute to the growing field of sports analytics.Methodology

```{=html}
<!-- -->
```

## Methodology 
-   Data Preparation:

    -   Filtered passing plays; cleaned data to 17,330 observations. Response variable: offensive yards gained (excluding penalties). Predictors: number of defenders in the box and pass rushers. Exploratory Data Analysis (EDA):

    -   Visual tools: boxplots, heatmaps, correlation matrix. Pass rushers showed slightly stronger correlation with play result (0.032) than defenders in the box (0.006).

    -   

-   Modeling:

    -   Linear Regression: Captures direct linear relationships.

    -   Generalized Additive Model (GAM): Accounts for non-linear relationships with smooth terms.

    -    Train-test split (80/20) and 5-fold cross-validation for performance evaluation.

    -   Metrics: RMSE, MAE, and R-squared. \## Results

```{r cars, echo = TRUE}
summary(cars)
```

## Included Plots

```{r pressure}
plot(pressure)
```

## Conclusion
